# vrnmf_germline

## Loading the data
The inference requires only a matrix of window-specific mutation rates. Below we use a matrix calculated based on TOPMed dataset:
```{r}
library(vrnmf)
library(spacemut)

rate.info <- read.table("http://pklab.med.harvard.edu/ruslan/spacemut/tracks/TOPMed_10kb.txt",header=TRUE)
```

```{r,results="hide"}
head(rate.info[,1:6])
```

```{r, echo=FALSE, results='asis'}
knitr::kable(head(rate.info[,1:6]))
```

Retain only columns of regional mutation rates:
```{r}
rate <- rate.info[,-c(1:3)]
```

## Vrnmf inference of spectra of mutational components

First, matrix is preprocessed and required statistics of covariance matrix are estimated using `vol_preprocess` routine:
```{r}
vol.info <- vol_preprocess(rate)
```


Volume-regularized NMF is applied to the co-occurence mutation matrix stored in the object `vol.info` to infer `n.comp = 14` components using volume weight `wvol = 7.9e-3` (see the manuscript for the motivation behind the choice of parameters). Alternating optimization was executed for `n.iter = 3e+3` iterations with `vol.iter` and `c.iter` updates of matrices inside each iteration:

```{r,results="hide"}
vr <- volnmf_main(vol.info, n.comp = 14, wvol = 7.9e-3, seed = 1,
                  n.iter = 3e+3, vol.iter = 2e+1, c.iter = 2e+1, 
                  verbose = FALSE)
```

Non-normalized spectra of mutational components are available in the matrix `vr$C`. To reflect the results in the paper, mutational spectra are then renormalized to the genome-average vector of mutation rates:

```{r}
S <- vr$C*vol.info$col.factors/colMeans(rate)
rownames(S) <- colnames(rate)
colnames(S) <- paste("comp.",1:ncol(S))
```

Note that order of components (columns of `S`) is arbitrary, and may not be consistent with that used in the manuscript. 
Spectra can now be visualized:
```{r,warning=FALSE, message=FALSE, fig.width = 4, fig.height = 4}
draw.signature(S[,2])
```

Additionally, we can explore reflection properties of components:
```{r,warning=FALSE, message=FALSE, fig.width = 4, fig.height = 4}
plot.reflection.matrix(S)
```

And visualize reflections of individual components:
```{r,results="hide",fig.width = 4, fig.height = 4}
reflection.scatter(2,3,S)
```

## Estimation of intensities of mutational components

Given known spectra of mutational components, intensities can then be obtained using non-negative least squares. A fast per-window estimation is implemented in a routine `infer_intensities`. However, we will additionally model offset spectrum, that would reflect all unaccounted mutational forces, using a routine `factor_intensities` (it takes a few to ten minutes to optimize):

```{r,results="hide"}
intensities.info <- factor_intensities(as.matrix(vr$C), t(as.matrix(vol.info$X.process)), 
                                  fit.nmf = FALSE, fit.factor = TRUE, 
                                  n.iter = 1e+2, verbose = FALSE) 
```

```{r}
str(intensities.info)
```

```{r}
intensities <- t(intensities.info$intensities)
```

Intensities can now be visualized:
```{r,results="hide",fig.width = 6, fig.height = 3}
plot.intensities(intensities[,13], rate.info, chr=16, start=0,end=4e+7,  span.wind=30)
```

